---
output: rmarkdown::github_document
always_allow_html: yes
bibliography: inst/references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  dpi = 300
)
```

# TreeTracer

`r emo::ji("tree")` `r emo::ji("tree")` `r emo::ji("tree")` `r emo::ji("tree")` `r emo::ji("tree")` `r emo::ji("tree")` `r emo::ji("tree")` `r emo::ji("tree")` `r emo::ji("tree")` `r emo::ji("tree")` `r emo::ji("tree")` `r emo::ji("tree")` `r emo::ji("tree")` `r emo::ji("tree")`

The beginnings...

TreeTracer is an R package for creating trace plots of trees from random forests fit using the randomForest R package. Trace plots are useful tools for visually comparing trees from a random forest. See @urbanek:2008 for additional information about trace plots. 

Examples:

```{r}
# Load the package
library(TreeTracer)
```

```{r}
# Fit a random forest using the iris data
set.seed(71)
iris.rf <- randomForest::randomForest(Species ~ ., data = iris)
```

```{r}
# Generate a trace plot of the first 25 trees in the forest
trace_plot(
  rf = iris.rf,
  train = iris[, -5],
  tree_ids = 1:10
)
```

```{r}
# Adjust the width of horizontal feature lines and alpha of traces
trace_plot(
  rf = iris.rf,
  train = iris[, -5],
  tree_ids = 1:10,
  width = 0.5,
  alpha = 0.25
)
```

## References

